Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
overdensity

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
set of spherical polygons to select batches of stars that have small latitude in
the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
\cite{Koposov:2010}.
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream system, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the over-density around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

As a first selection of GD-1 member stars, we select stars with proper motions
$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
1~\masyr$, as visualized by the rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars the pass the
selection, plotted in the GD-1 coordinate system.
From kinematic selection alone, the stream is already identifiable as the
over-density of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
\phi_1 \lesssim 0^\circ$.

@@

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp over-density in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material that has not yet phase-mixed in multiple dimensions.

We also measured the median proper motion profile along the stream (panels
(d)--(e)) and find that these are consistent with previously measured mean
proper motions from SDSS-USNOB (\citealt{Koposov:2010}).
Here, the proper motion gradient is resolved in striking detail for individual
stars from \gaia\ (see, e.g., elongated over-density in upper right panel of
\figurename~\ref{fig:selection}).

@@

To initialize the model stream generation, we assume that the present day (i.e.
final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
of the observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$
then forward-generate the model stream using the method discussed above, and
release stream particles at each timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
We therefore don't expect that the observed surface density variations could be
caused by, e.g., epicyclic over-densities (\citealt{Kupper:2012}).

@@

The central panel of Figure~\ref{fig:features} shows the map of GD-1, with fields of interest marked in shades of purple for spur-related fields, and shades of blue for blob-related fields.
In both cases, we analyze the stream region (darkest shade), overdensity off from the stream (medium shade) and use a field on the opposite side of the stream from the overdensity as a control (lightest shade).
The $\phi_1$ range is the same for all three fields associated with a given feature, and the control and feature field have the same area, while the stream field is smaller to avoid cross-contamination.

@@

\clearpage

>>>>

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
set of spherical polygons to select batches of stars that have small latitude in
the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
\cite{Koposov:2010}.
We convert the sky coordinates and proper motions of the parallax-selected stars
to the GD-1 stream system, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

As a first selection of GD-1 member stars, we select stars with proper motions
$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
1~\masyr$, as visualized by the rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars the pass the
selection, plotted in the GD-1 coordinate system.
From kinematic selection alone, the stream is already identifiable as the
overdensity of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
\phi_1 \lesssim 0^\circ$.

@@

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material that has not yet phase-mixed in multiple dimensions.

We also measured the median proper motion profile along the stream (panels
(d)--(e)) and find that these are consistent with previously measured mean
proper motions from SDSS-USNOB (\citealt{Koposov:2010}).
Here, the proper motion gradient is resolved in striking detail for individual
stars from \gaia\ (see, e.g., elongated overdensity in upper right panel of
\figurename~\ref{fig:selection}).

@@

To initialize the model stream generation, we assume that the present day (i.e.
final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
of the observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$
then forward-generate the model stream using the method discussed above, and
release stream particles at each timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
We therefore don't expect that the observed surface density variations could be
caused by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).

@@

The central panel of \figurename~\ref{fig:features} shows the map of GD-1, with fields of interest marked in shades of purple for spur-related fields, and shades of blue for blob-related fields.
In both cases, we analyze the stream region (darkest shade), overdensity off from the stream (medium shade) and use a field on the opposite side of the stream from the overdensity as a control (lightest shade).
The $\phi_1$ range is the same for all three fields associated with a given feature, and the control and feature field have the same area, while the stream field is smaller to avoid cross-contamination.

@@

% \clearpage
